## 0. Libraries

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
#install.packages("VIF")
#install.packages('caTools')
#devtools::install_github("sachsmc/plotROC")
#library(plotROC)
library(caTools)
library(purrr)
library(plyr)
library(dplyr)
library(ggplot2)
library(gridExtra)
#library(cowplot)
library(ROCR)
require(HH)
library(pscl)
library(caret)
library('ROSE')
library(ResourceSelection)
library(VIF) 
```

## 1. Know Your Data

Loading data:

```{r, include=FALSE}
loan_data <- read.csv("../data/loans.csv", header = T)
```

Exploring the variables:

```{r, include=FALSE}
summary(loan_data)
```

## 2. Loan Status Preparation

```{r, include=FALSE}

loan_data$loan_status[loan_data$status == 'Charged Off' || loan_data$status == 'Default'] <- 'Bad'

loan_data$loan_status[loan_data$status == 'Fully Paid'] <- 'Good'
loan_data$loan_status <- as.factor(loan_data$loan_status)

# print out results
loan_data %>% group_by(loan_status) %>% summarise(no_rows = length(loan_status))
```

## 3. Exclusion of Irrelevant Variables

```{r, include=FALSE}
loan_data <- loan_data[ , -which(names(loan_data) %in% c("loanID","employment", "state"))]
```

## 4. Transforming Data - Categories

```{r, echo=F, include=T, warning=F, message=F}
plot(loan_data$length, col="blue", main="Histogram of Employment Length - Before")
xfit<-seq(min(as.numeric(loan_data$length)),max(as.numeric(loan_data$length)),length=500) 
yfit<-dnorm(xfit,mean=mean(as.numeric(loan_data$length)),sd=sd(as.numeric(loan_data$length))) 
yfit <- yfit*diff(h$mids[1:2])*length(as.numeric(loan_data$length)) 
lines(xfit, yfit, col="black", lwd=2)
```

```{r, echo=F, include=T, warning=F, message=F}
from <- c('n/a', '< 1 year', '1 year', '2 years', '3 years', '4 years', '5 years',
          '6 years', '7 years', '8 years', '9 years', '10+ years')
to <- c("missing", '< 1 year', '1 to 3 years', '1 to 3 years', '1 to 3 years', '4 to 6 years', '4 to 6 years', '4 to 6 years', '7 to 9 years', '7 to 9 years', '7 to 9 years', '10+ years')

loan_data$length <- mapvalues(loan_data$length, from = from, to = to)
loan_data$length <- as.factor(loan_data$length)

#plot(loan_data$length, col="blue", main="Histogram of Employment Length")

hist(as.numeric(loan_data$length))
xfit<-seq(min(as.numeric(loan_data$length)),max(as.numeric(loan_data$length)),length=500) 
yfit<-dnorm(xfit,mean=mean(as.numeric(loan_data$length)),sd=sd(as.numeric(loan_data$length))) 
yfit <- yfit*diff(h$mids[1:2])*length(as.numeric(loan_data$length)) 
lines(xfit, yfit, col="black", lwd=2)

summary(loan_data$length)
```



#### Income

```{r include=FALSE}
plot(loan_data$income)

loan_data$income_cat[ loan_data$income < 9700 ] <- "< 9700"
loan_data$income_cat[ loan_data$income >= 9700 & loan_data$income <= 39475 ] <- "9700 - 39475"
loan_data$income_cat[ loan_data$income > 39475 & loan_data$income <= 84200 ] <- "39476 - 84200"
loan_data$income_cat[ loan_data$income > 84200 & loan_data$income <= 160725 ] <- "84200 - 160725"
loan_data$income_cat[ loan_data$income > 160726 & loan_data$income <= 204101 ] <- "160726 - 204101"
loan_data$income_cat[ loan_data$income > 204101 & loan_data$income <= 510300 ] <- "204102 - 510300"
loan_data$income_cat[ loan_data$income > 510300 ] <- "> 510300"

loan_data$income_cat <- as.factor(loan_data$income_cat)

summary(loan_data$income_cat)

catresult <- c( 1, 11, 7400, 27405, 13071, 1124, 926, 62)
plot(catresult)

```




## 5. Transforming Data - Removing Records

```{r, include=FALSE}
loan_data <- loan_data[!(loan_data$status %in% c('Current', 'Late (16-30 days)', 'Late (31-120 days)')),]

loan_data <- loan_data[!is.na(loan_data$amount), ]
loan_data <- loan_data[!is.na(loan_data$rate), ]
loan_data <- loan_data[!is.na(loan_data$payment), ]
loan_data <- loan_data[!is.na(loan_data$income), ]
loan_data <- loan_data[!is.na(loan_data$length), ]

summary(loan_data)
```

Credit Ratios:

```{r include=F}
m <- summary(loan_data$revolRatio)[[3]]

loan_data <- (loan_data %>% mutate(
  revolRatio = coalesce(revolRatio, m)
))

m <- as.integer(summary(loan_data$bcOpen)[[3]])

loan_data <- (loan_data %>% mutate(
  bcOpen = coalesce(bcOpen, m)
))

m <- as.double(summary(loan_data$bcRatio)[[3]])

loan_data <- (loan_data %>% mutate(
  bcRatio = coalesce(bcRatio, m)
))

summary(loan_data)
```


## 6. Variable Selection - Correlation

```{r echo=FALSE, INCLUDE=FALSE}
cor(loan_data[, sapply(loan_data, class) != 'factor' ]) # factors can't be in correlation
```

## 7. Variable Selection - Income

One major contributing factor to the ability to pay off loans is to have an income. 

```{r echo=F}
hist(loan_data$income, main="Figure 1 - Initial Income Histogram", xlab="Income")
```

The income histogram immediately depicts that there are some extreme values, making this data appear right-skewed. The Z-Score's for the income field depict that the median is -0.1604 standard deviations away from the mean, but the maximum value is 115.5 standard deviations; suggesting we have some outliers to either: remove, inferr, or possibly turn income into a categorical variable depicting income range.

```{r echo=F}
loan_sd <- sd(loan_data$income)*sqrt((length(loan_data$income)-1)/(length(loan_data$income)))
loan_data$income_zscore <- (loan_data$income - mean(loan_data$income)) / loan_sd

summary(loan_data$income_zscore)
```


```{r echo=F}
loan_data$income[ loan_data$income_zscore > 4 ] <- median(loan_data$income, na.rm=TRUE)
loan_data$income[ loan_data$income_zscore < -4 ] <- median(loan_data$income, na.rm=TRUE)

hist(loan_data$income, main="Figure 2 - Income within +- 4 Std Deviations", xlab="Income")
boxplot(loan_data$income, main="Figure 3 - Income within +- 4 Std Deviations",  ylab="Income")
```

Omitting 134 containing income within +- 4 standard deviations from the mean and converting them to NA values, we can see that our income sample is right-skewed. In attempt to normalize the data and reduce the extremeness of the tail, depicted in the boxplot.

```{r echo=F}
loan_data$income_sqrt <- sqrt(loan_data$income)
hist(loan_data$income_sqrt, main="Figure 4 - Sqrt of Income", xlab="Sqrt - Income")
```


## 8. The Logistic Model

After reviewing the data and identifying correlations between variables, the following independent variables were selected to determine our dependent variable loan_status: debtIncRat, pubRec, income, totalAcc, home, length, delinq2yr, rate, grade, inq6mth, reason, amount, totalRevBal, revolRatio. Due to the correlation between amount and payment, this was excluded. There is a slight correlation between payment, income and amount, however not a strong enough correlation to conclude that they should be excluded.

```{r echo=F, INCLUDE=F}
set.seed(312)
#names(subset(loan_data))

loan_data.selected_fields <- subset(loan_data, select = c(1, 3, 5:8, 11:14, 16:18, 27, 30))
sample_data <- sample.split(loan_data.selected_fields$loan_status, 0.80)
loan_data.training.data <- subset(loan_data.selected_fields, sample_data == TRUE)
loan_data.test.data <- subset(loan_data.selected_fields, sample_data == FALSE)

loan_data.model.full <- glm(loan_status ~ ., data = loan_data.training.data, family = "binomial")
summary(loan_data.model.full)
```

The prediction was ran against the fitted data with a threshold of 0.5, where any value >= 0.5 is a Good loan, and anything < 0.5 is bad.

```{r echo=F}
threshold = 0.5
prob_full <- predict(loan_data.model.full, newdata = loan_data.test.data, type = "response")
prob_cut <- ifelse(prob_full > threshold, 'Good', 'Bad')

cTab <- table(prob_cut, loan_data.test.data$loan_status)
v <- sum(diag(cTab)) / sum(cTab)

print(paste('Proportion correctly predicted = ', v))
```

According to our training sample, our model predicts 77.66% of the data, suggesting we have a decent model fitting our data.

```{r echo=F}
vif(loan_data.model.full)
```